---
title: "Solutions Chapter 1"
author: "Michael Mayer"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
    df_print: paged
    theme: paper
    code_folding: show
    math_method: katex
editor_options: 
  chunk_output_type: console
knit: (function(input, ...) {rmarkdown::render(input, output_dir = "../docs")})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE
)
```

# Exercise on linear regression

```{r}
library(tidyverse)

diamonds <- mutate_if(diamonds, is.ordered, factor, ordered = FALSE)

fit <- lm(price ~ carat + color + cut + clarity, data = diamonds)
summary(fit)
```

**Comments**

- **Model quality:** About 92% of price variations are explained by covariates. Typical prediction error is 1157 USD.
- **Effects:** All effects point into the intuitively right direction (larger stones are more expensive, worse color are less expensive etc.)
- **Practical perspective:** Additivity in color, cut and clarity are not making sense. Their effects should get larger with larger diamond size. This can be solved by adding interaction terms with carat or, much easier, to switch to a logarithmic response.
 
# Exercise on GLMs

```{r}
library(tidyverse)

diamonds <- mutate_if(diamonds, is.ordered, factor, ordered = FALSE)

fit <- glm(
  price ~ log(carat) + color + cut + clarity, 
  data = diamonds, 
  family = Gamma(link = "log")
)
summary(fit)

mean(predict(fit, type = "response")) / mean(diamonds$price) - 1
```

**Comment:** The coefficients are very similar to the linear regression with log(price) as response. This makes sense as we interpret the coefficients in the same way! The bias is small, but not exactly 0 (because log is not the natural link of the Gamma GLM).
